---
title: "1. Summary"
output:
  html_document:
    code_folding: "hide"
    theme: paper
---

```{r setup, include=FALSE}
# rmarkdown::render_site()
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F)
library(kableExtra)
library(formattable)
library(eutopsQC)
path <- log

rm_names <- unique(c(chrY_names,non_CpG_names,snp_names,zhou_list))
```

This page provides an overview of the sample failure rate per beadchip. Of note, `r length(rm_names)` probes (~10% of probes on the EPIC array) are removed during QC as default as they have been flagged as Chromosome Y probes, non-CpG probes, SNP probes, or otherwise unreliable probes, leaving a total of `r as.character(866238-length(rm_names))` probes. The remainder of probes may be removed during QC for other reasons (e.g. failed detection p value).

```{r}
# Construct data frame
n = 1
log_df <- data.frame(Experiment = run.name,
                     Sample_count = rep(NA,n),
                     Probe_count = rep(nrow(beta),n),
                     Removed_sample_count = rep(NA,n))

log_df$Probe_count <- paste0(nrow(beta), " / ",
                             log_data$n_probes[1],
                             " (", round(nrow(beta)/log_data$n_probes[1]*100, 1), "%)")

log_df$Sample_count <- ncol(beta)
sentrix_id <- stringr::str_split(b, "_", simplify = T)[,1]

if(!is.na(log_data$rm_sample_list)){
  log_df$Removed_sample_count <- paste0(length(log_data$rm_sample_list))
} else {
  log_df$Removed_sample_count <- "0"
}

colnames(log_df) <- c('Experiment',
                      'Samples',
                      'Probes',
                      'Removed samples')

kable(log_df) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## List of removed samples

```{r}
if((length(log_data$rm_sample_list)>0 & !is.na(paste0(log_data$rm_sample_list, collapse = ",")))){
rm <- pheno[match(na.omit(log_data$rm_sample_list), pheno$basename),] |> 
  dplyr::distinct()

rm |> 
  dplyr::filter(!is.na(basename)) |> 
  DT::datatable()
} else {
  cat("No samples were removed.")
}
```

