---
title: "1. Beadchip summary"
output:
  html_document:
    code_folding: "hide"
    theme: paper
---

```{r setup, include=FALSE}
# rmarkdown::render_site()
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F)
library(kableExtra)
library(formattable)
library(eutopsQC)
path <- log

rm_names <- unique(c(chrY_names,non_CpG_names,snp_names,zhou_list))
```

This page provides an overview of the sample failure rate per beadchip. Of note, `r length(rm_names)` probes (~10% of probes on the EPIC array) are removed during QC as default as they have been flagged as Chromosome Y probes, non-CpG probes, SNP probes, or otherwise unreliable probes, leaving a total of `r as.character(866238-length(rm_names))` probes. The remainder of probes may be removed during QC for other reasons (e.g. failed detection p value).

```{r}
# Scan for beadchips in PATH
beadchips <- unique(stringr::str_split(pheno$basename, "_", simplify = T)[,1])

# Construct data frame
n <- length(beadchips)
log_df <- data.frame(Beadchip = beadchips,
                     Sample_count = rep(NA,n),
                     Probe_count = rep(nrow(beta_merged),n),
                     Removed_sample_count = rep(NA,n))

files <- list.files(path, full.names = T)

if(sum(grepl('_log_data.Rdata',files))==1){
  load(files[grepl('_log_data.Rdata',files)])
  } else {
  files <- files[grepl('_log_data.Rdata',files)]
  
  for (f in files){
    load(f)
    
    if(f == files[1]){
    log_data_tmp <- log_data
    } else {
      
      for (x in 1:length(log_data_tmp)){
        log_data_tmp[[x]] <- c(log_data_tmp[[x]], log_data[[x]])
      }
    }
    
    log_data <- log_data_tmp
  }
  
}


log_df$Probe_count <- paste0(nrow(beta_merged), " / ",
                             log_data$n_probes[1],
                             " (", round(nrow(beta_merged)/log_data$n_probes[1]*100, 1), "%)")

b <- pheno$basename
sentrix_id <- stringr::str_split(b, "_", simplify = T)[,1]

for (i in 1:nrow(log_df)){
  
  if(any(grepl(log_df$Beadchip[i], log_data$rm_sample_list))){
    tmp <- log_data$rm_sample_list[grepl(log_df$Beadchip[i], log_data$rm_sample_list)]
    log_df$Sample_count[i] <- sum(sentrix_id == log_df$Beadchip[i])-length(tmp)
    log_df$Removed_sample_count[i] <- paste0(length(tmp), " (",
                                          length(tmp)/8*100, "%)")
    } else {
    log_df$Sample_count[i] <- sum(sentrix_id == log_df$Beadchip[i])
    log_df$Removed_sample_count[i] <- "0 (0%)"
  }
  
}

colnames(log_df) <- c('Beadchip',
                      'Samples',
                      'Probes',
                      'Removed samples')

# kable(log_df)%>%
#   kable_styling(bootstrap_options = c("striped", "hover"))

x <- log_df |>
  formattable(list(
              Beadchip = formatter("span",
                                   style = x ~ style(color = ifelse(log_df$`Removed samples` == "0 (0%)", "seagreen", "palevioletred"))),
              Samples = formatter("span",
                                           x ~ icontext(ifelse(log_df$`Removed samples` == "0 (0%)", "ok", "remove"),
                                                        x),
                                           style = x ~ style(color = ifelse(log_df$`Removed samples` == "0 (0%)", "seagreen", "palevioletred"))),
              Probes = formatter("span",
                                 style = x ~ style(color = ifelse(readr::parse_number(stringr::str_extract_all(log_df$Probes, "\\([^()]+\\)")[[1]]) > 80, "seagreen",
                                                          "palevioletred")))))

as.datatable(x, rownames = FALSE,
             options = list(pageLength = 10))
```

## List of removed samples

```{r}
if((length(log_data$rm_sample_list)>0 & !is.na(paste0(log_data$rm_sample_list, collapse = ",")))){
rm <- pheno[match(na.omit(log_data$rm_sample_list), pheno$basename),] |> 
  dplyr::distinct()

rm |> 
  dplyr::filter(!is.na(basename)) |> 
  DT::datatable()
} else {
  cat("No samples were removed.")
}
```

