---
title: "7. Dimensionality reduction"
output:
  html_document:
    code_folding: "hide"
    theme: paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(plotly)
library(dplyr)
library(ggplot2)
library(uwot)

coerce_numeric <- function(beta_merged){
    beta_merged_tmp <- matrix(NA, nrow=nrow(beta_merged), ncol=ncol(beta_merged))
    for(i in 1:ncol(beta_merged)){
      beta_merged_tmp[,i] <- as.numeric(beta_merged[,i])
    }
    colnames(beta_merged_tmp) <- colnames(beta_merged)
    rownames(beta_merged_tmp) <- rownames(beta_merged)
    beta_merged <- beta_merged_tmp
    rm(beta_merged_tmp);gc()
    return(beta_merged)
}

# fix any potential difftime issues
pheno <- pheno |> 
  dplyr::mutate_if(is.difftime, as.numeric)
```

Two methods (linear and non-linear) are used for dimensionality reduction, which both have advantages and disadvantages:

1. UMAP (Uniform manifold approximation and projection): non-linear dimensionality reduction
2. Multidimensional scaling (principal component analysis): linear dimensionality reduction

A heatmap of p values for principal components and (if provided, their association with provided phenotypic values) is shown on the bottom of the page to identify potential batch effects.

# UMAP

```{r fig.align = "center", fig.width = 4.5, fig.height = 6}
if(ncol(beta_merged)<=15){
  n <- ncol(beta_merged)
} else {
 n <- 15 
}

beta_merged <- coerce_numeric(beta_merged)

intersect <- intersect(pheno$basename, colnames(beta_merged))
beta_merged <- beta_merged[,match(intersect, colnames(beta_merged))]
pheno <- pheno[match(intersect, pheno$basename),]

# Grab top 20000 variable CpGs
x <- matrixStats::rowSds(beta_merged)
names(x) <- rownames(beta_merged)
top10 <- x[order(x, decreasing = T)][1:30000]
b <- beta_merged[names(top10),]


umap <- uwot::umap(t(b), n_neighbors = n)
pheno$umap1 <- umap[,1]
pheno$umap2 <- umap[,2]

variables <- colnames(pheno)[!colnames(pheno) %in% c("basename",
                                                     "umap1",
                                                     "umap2", "widage")]

plt <- htmltools::tagList()
i <- 1
for (var in variables){
  x <- pheno %>%
    ggplot(aes(x = umap1,
               y= umap2,
               colour = .data[[var]])) +
    geom_point() +
    ggtitle(var) +
    theme(legend.title = element_blank(),
          aspect.ratio = 1)
  
  # Print an interactive plot
  # Add to list
  plt[[i]] <- as_widget(ggplotly(x))
  i <- i + 1
}


plt
```

# MDS plots

```{r fig.align = "center", fig.width = 4.5, fig.height = 4.5}
d <- dist(t(b))
fit <- cmdscale(d,eig=TRUE, k=2)

x <- fit$points[,1]
y <- fit$points[,2]

pheno$pc1 <- x
pheno$pc2 <- y

# Save pheno
save(pheno, file = paste0(log, "/pheno_qc.Rdata"))

variables <- colnames(pheno)[!colnames(pheno) %in% c("basename",
                                                     "umap1",
                                                     "umap2",
                                                     "pc1", "pc2", "widage")]
plt <- htmltools::tagList()
i <- 1
for (var in variables){
  x <- pheno %>%
    ggplot(aes(x = pc1,
               y = pc2,
               colour = .data[[var]])) +
    geom_point() +
    ggtitle(var) +
    theme(legend.title = element_blank(),
          aspect.ratio = 1)
  
  plt[[i]] <- as_widget(ggplotly(x))
  i <- i + 1
}

plt
```


# PCA

```{r}
fit <- prcomp(t(b[1:30000,]),retx=TRUE,scale.=FALSE,center=TRUE)

if(ncol(b)<=10){
  P <- ncol(b)-1
  } else {
    P <- 10
} # gives first P principal components

pdat <- pheno %>%
  dplyr::select(where(~ length(unique(na.omit(.)))>1)) |>  # Keep only columns which have more than one unique value for batch effect analysis.
  dplyr::select(-c(basename, umap1, umap2, pc1, pc2)) |> 
  mutate_if(is.character,as.factor) |> 
  as.data.frame()

pval <- matrix(NA, nrow = ncol(pdat), ncol = P)
X <- fit$x[,1:P]

rownames(pval) <- colnames(pdat)
colnames(pval) <- colnames(X)

# loop over each phenotypic trait
for (i in 1:ncol(pdat)){
  
  #loop over each of the P principal components
  for (j in 1:P){
    
    # if factor do kruskal wallis test
    if (is.factor(pdat[,i])){
      tst <- kruskal.test(X[,j],pdat[,i])
      pval[i,j] <- tst$p.value
    }
    
    # if numeric look for correlation
    if (is.numeric(pdat[,i])){
      tst <- cor.test(X[,j],pdat[,i])
      pval[i,j] <- tst$p.value
    }
  }
}

pval[pval==0] <- 1e-100

plot_ly(x=colnames(pval),
        y=rownames(pval),
        z=log10(pval),
        colors = colorRamp(c("red", "yellow", "white")),
        type='heatmap')%>%
  config(displayModeBar = FALSE)
```